\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-14 paper.tex
Fix whitespace

\linenumbers
\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely,ignatieva2023distribution}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, an ARG  has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the outcome of such a stochastic process or a
genetic genealogy inferred from a sample of genome sequences.
However, the distinction between process and data structure is not clear cut
and subfields use the term
differently (see Appendix XXX).

@@

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we define the genome ARG (gARG) and contrast it with the event ARGs (eARG).
% Retrospective vs prospective
% TODO: This needs revision - see also https://github.com/tskit-dev/what-is-an-arg-paper/pull/380
We continue with a description of the usually held retrospective view of ARGs, but note that ARGs can also be viewed as prospective, particularly in forwards-time simulations. In the context of forwards-time simulations, we mention the simplification operation, which removes ancestral genome material that was not inherited.
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARG and gARG, which opens a rich set of details about ARGs and local trees, local unary nodes, and levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important implications for computational efficiency, which is critical if we want to use ARGs for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE? Finally, the literature on ARGs and associated aspects is vast, particularly due to different definitions and implementations. We cover some important aspects in appendices, including a brief history of ancestral graphs (Appendix XXX), description of the big and little ARGs (Appendix ZZZ), survey of ARG inference methods (Appendix YYY), and ARGs at an individual vs cell lineage level (Appendix WWW).

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We are not concerned here with mutational processes or observed sequences,
but consider only processes of inheritance,
following the standard practice in coalescent theory.

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, and $E$ be the set of edges.
We will usually define a set of nodes $S \subseteq N$
representing the sampled genomes
(but see section XXX for some subtlety around this point).
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is ``annotated'' with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes can respectively represent ancestor and descendant genomes,
which can be separated by multiple generations.
We will use these two sets of terms interchangeably.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. Once the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example genome ARG (gARG) embedded in a pedigree.
(A) Diploid individuals (blue), visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events (red) between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown with shaded colour.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes contained within a parent. In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
% Following the above definitions, corresponding data for this gARG is shown in Supplementary? Table XXX.

These definitions follow naturally from the outcome of events
associated with the inheritance of DNA between generations.
As such,
there are some important consequences to these definitions.
Firstly, there is
no limitation on a mating system, ploidy, or age structure in a gARG.
By making the
basic unit an individual's genomes,
any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
species) can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance can be directly expressed, such as
gene conversions,
multiple simultaneous recombinations,
and so on.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity (number of child nodes), including ``unary''
nodes with a single child (see Section XXX).
% TODO should rephrase this from parent/child to ancestor/descendant?
Fourth, recombination is modelled in terms of
its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
parental nodes from which this child has inherited genome segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the processes of genetic inheritance
with recombination,
and is agnostic to the details of those processes.

% Firstly, a gARG may contain nodes that are both samples \emph{and}
% internal nodes in the local trees.
% This ability is useful when we have pedigree information
% along with genetic data
% \cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
% where we may have many generations of internal sample nodes
% whose genomes have been sequenced.
% Another situation in which it is important that
% we do not assume samples are all contemporary ``leaf'' nodes
% (or that all internal nodes are either coalescences or recombinations)
% is when we are incorporating ancient genomes into ARG
% inference~\citep{speidel2021inferring,wohns2022unified}.
% Similarly, a gARG may contain nodes that
% do not correspond to any particular event in the ancestry of the samples.
% For example, node \textsf{h} is the direct ancestor of \textsf{d} in
% Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
% coalescence or a recombination. Such nodes would usually be removed
% from the graph so that \textsf{d} descends directly from \textsf{k} (see
% the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
% a representational perspective.
% Secondly, because we do not classify nodes by event type
% (common ancestor or recombination) there is no limitation
% on the complexity that can be modelled---multiple recombinations
% and coalescences can happen simultaneously in the same genome,
% a common occurance in large deeply sequenced
% pedigrees~\citep{hayes20191000,RosFreixedes2022}.
% Finally, there is a subtle and important point about how recombination
% is represented: which genome in the organismal lifecycle
% do nodes represent?
% In principle, we can interpret nodes as representing genomes
% at any point in the life cycle [supplementary Figure ?]. Hudson
% views nodes as representing gametes~\citep{hudson1983properties}.
% However, we argue that the most intuitive
% approach is to let nodes represent the genomes present in diploid
% individuals in each generation, as in
% Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
% is not represented by a single ``recombination node'' (as in an eARG), but with
% two parent genomes between which recombination takes place
% (e.g.\ nodes \textsf{e} and \textsf{f}), and a
% single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
% edges that link this recombinant with its two parents carry the information
% concerning breakpoint location that would be associated with a
% recombination node in an eARG. [TODO finish up here with more
% clarity on what this means in terms of the required specifity
% of recombination events, and put in a forward ref to the simplification
% section where we talk about stacking up recombination events
% and their identifiability.]

\section*{Event ARGs}\label{eARG}
Genome ARGs are defined by
describing the details of genetic inheritance
between ancestor and descendant genomes.
In contrast, ARGs are classically defined
by describing the \emph{events} in the history of a sample of genomes.
The difference between gARGs and the prevalent views of eARGs
% what defines an ARG
are subtle and difficult to distinguish.
It is therefore useful to first survey some representative
recent definitions of
an ARG from the literature, to allow us to capture their common features
(see Appendix XXX for a full review).

@@

% NOTE: important to emphasise TWO things here, events and vagueness
% actual encoding mechanism (i.e. how trees fit into the ARG)
A common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which local trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occurred,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is usually assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event,
% Trying to emphasise here that this is just an example data encoding.
% I can imagine
% people nitpicking about how this specific way of doing it is bad
% for reason X.
along with an example encoding of the data.
There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Fig.~\ref{fig-event-arg}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined in the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
% JK: not sure what you're getting at here Gregor? The ordering matters
% because you have to write each event down precisely.
% Gregor: Importantly, the order of events matters to WHAT!?

@@

\foreach \u/\lab in {
        s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$,
        r0/$\textsf{a}$, r1/$\textsf{b}$, r2/$\textsf{c}$,
        l0/$\textsf{a}$, l1/$\textsf{b}$, l2/$\textsf{c}$}
    \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};

\foreach \u/\lab in {
        s6/$\textsf{g}$,
        l4/$\textsf{g}$,
        r4/$\textsf{g}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

\node [nodelabel,anchor=north west] at ($(s3) + (\x1 + 0,0)$) {$x$};
\foreach \u/\lab in {
        s4/$\textsf{e}$,
        l3/$\textsf{e}$}
    \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {
        s5/$\textsf{f}$,
        r3/$\textsf{f}$}
    \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

@@

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG. (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see section XXX for a discussion
of this important point).
(C) A concrete encoding of the information (right).
Recombination nodes have a non-null breakpoint
and two parents.
}
\end{figure}

Whether we regard nodes as events or genomes is something of a
philosophical question, and in most senses the two are equivalent
(but see the discussion on ``unary nodes'' in section XXX).
From a concrete data-structure encoding perspective, the only real
difference between an eARG and a gARG is that in an eARG we store recombination
\emph{breakpoints} at nodes of out-degree two (recombination nodes)
whereas in a gARG we store \emph{inheritance intervals} on edges.
A key advantage of associating inheritance intervals with
edges in a gARG is that it allows us to describe genetic inheritance events other
than simple crossovers, such as multiple recombinations and
gene conversions. By explicitly stating the exact regions of genome
that are transmitted along an edge, we generalise the representation
to encompass any process that leads to the differential inheritance of
genetic material along the genome.

>>>>

\linenumbers
\section{Introduction}
\label{sec-intro}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely,ignatieva2023distribution}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, an ARG  has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the outcome of such a stochastic process or a
genetic genealogy inferred from a sample of genome sequences.
However, the distinction between process and data structure is not clear cut
and subfields use the term
differently (see Appendix~\ref{sec-arg-history}).

@@

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we define the genome ARG (gARG) and contrast it with the event ARGs (eARG).
% Retrospective vs prospective
% TODO: This needs revision - see also https://github.com/tskit-dev/what-is-an-arg-paper/pull/380
We continue with a description of the usually held retrospective view of ARGs,
but note that ARGs can also be viewed as prospective, particularly in forwards-time
simulations. In the context of forwards-time simulations, we mention the
simplification operation, which removes ancestral genome material that was not
inherited.
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARG and gARG,
which opens a rich set of details about ARGs and local trees, local unary nodes, and
levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision
about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important
implications for computational efficiency, which is critical if we want to use ARGs
for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
Finally, the literature on ARGs and associated aspects is vast, particularly
due to different definitions and implementations. We cover some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

\section{Genome ARGs}
\label{sec-gARG}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We are not concerned here with mutational processes or observed sequences,
but consider only processes of inheritance,
following the standard practice in coalescent theory.

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, and $E$ be the set of edges.
We will usually define a set of nodes $S \subseteq N$
representing the sampled genomes
(but see Section~\ref{sec-retro-pro} for some subtlety around this point).
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is ``annotated'' with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes can respectively represent ancestor and descendant genomes,
which can be separated by multiple generations.
We will use these two sets of terms interchangeably.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. Once the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example genome ARG (gARG) embedded in a pedigree.
(A) Diploid individuals (blue), visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \noderef{a} to \noderef{p}. Black lines show inheritance paths connecting
genomes in the current generation (\noderef{a} to \noderef{d}) with their ancestors.
Genomes \noderef{a} and \noderef{c} are the product of two independent
recombination events (red) between
the paternal genomes \noderef{e}
and \noderef{f}, and regions of genome inherited are shown with shaded colour.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes contained within a parent. In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
Following the above definitions, corresponding data encoding for
this gARG is shown in Tab.~\ref{tab-gARG-data} (Appendix~\ref{sec-gARG-data}).

These definitions follow naturally from the outcome of events
associated with the inheritance of DNA between generations.
As such,
there are some important consequences to these definitions.
Firstly, there is
no limitation on a mating system, ploidy, or age structure in a gARG.
By making the
basic unit an individual's genomes,
any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
species) can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance can be directly expressed, such as
gene conversions,
multiple simultaneous recombinations,
and so on.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity (number of child nodes), including ``unary''
nodes with a single child (see Section~\ref{sec-locally-unary-edges}).
% TODO should rephrase this from parent/child to ancestor/descendant?
Fourth, recombination is modelled in terms of
its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see Section~\ref{sec-ARG-simplification})
% TODO: check above if we want to refer to sec-ARG-simplification or some other!
parental nodes from which this child has inherited genome segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the processes of genetic inheritance
with recombination,
and is agnostic to the details of those processes.

% Firstly, a gARG may contain nodes that are both samples \emph{and}
% internal nodes in the local trees.
% This ability is useful when we have pedigree information
% along with genetic data
% \cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
% where we may have many generations of internal sample nodes
% whose genomes have been sequenced.
% Another situation in which it is important that
% we do not assume samples are all contemporary ``leaf'' nodes
% (or that all internal nodes are either coalescences or recombinations)
% is when we are incorporating ancient genomes into ARG
% inference~\citep{speidel2021inferring,wohns2022unified}.
% Similarly, a gARG may contain nodes that
% do not correspond to any particular event in the ancestry of the samples.
% For example, node \noderef{h} is the direct ancestor of \noderef{d} in
% Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
% coalescence or a recombination. Such nodes would usually be removed
% from the graph so that \noderef{d} descends directly from \noderef{k} (see
% the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
% a representational perspective.
% Secondly, because we do not classify nodes by event type
% (common ancestor or recombination) there is no limitation
% on the complexity that can be modelled---multiple recombinations
% and coalescences can happen simultaneously in the same genome,
% a common occurance in large deeply sequenced
% pedigrees~\citep{hayes20191000,RosFreixedes2022}.
% Finally, there is a subtle and important point about how recombination
% is represented: which genome in the organismal lifecycle
% do nodes represent?
% In principle, we can interpret nodes as representing genomes
% at any point in the life cycle [supplementary Figure ?]. Hudson
% views nodes as representing gametes~\citep{hudson1983properties}.
% However, we argue that the most intuitive
% approach is to let nodes represent the genomes present in diploid
% individuals in each generation, as in
% Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
% is not represented by a single ``recombination node'' (as in an eARG), but with
% two parent genomes between which recombination takes place
% (e.g.\ nodes \noderef{e} and \noderef{f}), and a
% single resulting ``recombinant'' genome (e.g.\ \noderef{c}). The
% edges that link this recombinant with its two parents carry the information
% concerning breakpoint location that would be associated with a
% recombination node in an eARG. [TODO finish up here with more
% clarity on what this means in terms of the required specifity
% of recombination events, and put in a forward ref to the simplification
% section where we talk about stacking up recombination events
% and their identifiability.]

\section{Event ARGs}
\label{sec-eARG}
Genome ARGs are defined by
describing the details of genetic inheritance
between ancestor and descendant genomes.
In contrast, ARGs are classically defined
by describing the \emph{events} in the history of a sample of genomes.
The difference between gARGs and the prevalent views of eARGs
% what defines an ARG
are subtle and difficult to distinguish.
It is therefore useful to first survey some representative
recent definitions of
an ARG from the literature, to allow us to capture their common features
(see Appendix~\ref{sec-arg-history} for a full review).

@@

% NOTE: important to emphasise TWO things here, events and vagueness
% actual encoding mechanism (i.e. how trees fit into the ARG)
A common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which local trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occurred,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' flows along
the edges of the ARG (see Section~\ref{sec-eARG-to-gARG}).
% TODO: check above if we want to refer to sec-eARG-to-gARG or some other!
This is usually assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event,
% Trying to emphasise here that this is just an example data encoding.
% I can imagine
% people nitpicking about how this specific way of doing it is bad
% for reason X.
along with an example encoding of the data.
There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Fig.~\ref{fig-event-arg}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined in the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
% JK: not sure what you're getting at here Gregor? The ordering matters
% because you have to write each event down precisely.
% Gregor: Importantly, the order of events matters to WHAT!?

@@

\foreach \u/\lab in {
        s0/$\noderef{a}$, s1/$\noderef{b}$, s2/$\noderef{c}$,
        r0/$\noderef{a}$, r1/$\noderef{b}$, r2/$\noderef{c}$,
        l0/$\noderef{a}$, l1/$\noderef{b}$, l2/$\noderef{c}$}
    \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};

\foreach \u/\lab in {
        s6/$\noderef{g}$,
        l4/$\noderef{g}$,
        r4/$\noderef{g}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

\node [nodelabel,anchor=north west] at ($(s3) + (\x1 + 0,0)$) {$x$};
\foreach \u/\lab in {
        s4/$\noderef{e}$,
        l3/$\noderef{e}$}
    \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {
        s5/$\noderef{f}$,
        r3/$\noderef{f}$}
    \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\noderef{d}$, s6/$\noderef{g}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

@@

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG. (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
for a discussion of this important point).
(C) A concrete encoding of the information.
Recombination nodes have a non-null breakpoint
and two parents.
}
\end{figure}

Whether we regard nodes as events or genomes is something of a
philosophical question, and in most senses the two are equivalent
(but see the discussion on ``unary nodes'' in Section~\ref{sec-locally-unary-edges}).
From a concrete data-structure encoding perspective, the only real
difference between an eARG and a gARG is that in an eARG we store recombination
\emph{breakpoints} at nodes of out-degree two (recombination nodes)
whereas in a gARG we store \emph{inheritance intervals} on edges.
A key advantage of associating inheritance intervals with
edges in a gARG is that it allows us to describe genetic inheritance events other
than simple crossovers, such as multiple recombinations and
gene conversions. By explicitly stating the exact regions of genome
that are transmitted along an edge, we generalise the representation
to encompass any process that leads to the differential inheritance of
genetic material along the genome.

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks.
The vitally important and fundamental distinction between
a gARG and an eARG is that gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, which cannot be done
by specifying recombination events and breakpoints.
% FIXME this sentence is problematic. We should recast to talk about
% the outcomes of events, not talk about simulateousness (which will
% confuse people
Thus, we have the means to encode the \emph{outcome} of
multiple genetic inheritance events, and to
systematically describe the uncertainty around their
ordering (see Section~\ref{sec-precision}).
As we describe in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has significant advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.
